---
title: "Multi-stream monitoring"
author: ""
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-stream monitoring}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, results="hide", message=FALSE, warning=FALSE}
library(bycatch)
set.seed(123)
```

### Overview

In many fisheries, bycatch is monitored through multiple independent data streams. For example:

- **Observer-only coverage**: Traditional human observers on vessels
- **Electronic Monitoring (EM) only**: Video monitoring without observers
- **Both Observer and EM**: Vessels with both monitoring types simultaneously

The `bycatch` package supports multi-stream monitoring, where data from different monitoring programs are kept separate in the statistical model but share the same underlying bycatch rate. The detection rate of each stream is currently assumed to be 100%, allowing each stream to be added as a new likelihood component. 

In the special cases where the family is Poisson, data from different streams may be summed into a single stream (but for interpretation, keeping these separate may be easier).

### Simple two-stream example

Let's start with a basic example where we have both observer and EM coverage:

```{r data}
d <- data.frame(
  Year = 2010:2019,
  
  # Observer stream
  Takes_obs = c(2, 1, 3, 0, 2, 1, 0, 2, 1, 3),
  Sets_obs = rep(100, 10),    # 100 sets observed
  CovRate_obs = rep(20, 10),  # 20% coverage
  
  # EM stream  
  Takes_em = c(1, 0, 2, 1, 1, 0, 2, 1, 0, 1),
  Sets_em = rep(80, 10),      # 80 sets monitored by EM
  CovRate_em = rep(16, 10),   # 16% coverage
  
  # Total fishery effort (optional - for backward compatibility)
  Sets_total = rep(500, 10)   # 500 total sets per year
)
```

### Fitting a multi-stream model

To activate multi-stream mode, provide column names for the additional monitoring streams:

```{r, results="hide", message=FALSE, warning=FALSE}
fit <- fit_bycatch(
  Takes_obs ~ 1,             # Formula uses the observer stream
  data = d,
  time = "Year",
  effort = "Sets_obs",       # Observer effort
  covrate_obs = "CovRate_obs",  # Observer coverage rate
  takes_em = "Takes_em",     # EM takes (activates multi-stream)
  effort_em = "Sets_em",     # EM effort
  covrate_em = "CovRate_em",    # EM coverage rate
  family = "poisson",
  time_varying = FALSE
)

# OLD APPROACH: Using effort_total (still works)
# fit <- fit_bycatch(
#   Takes_obs ~ 1,
#   data = d,
#   time = "Year",
#   effort = "Sets_obs",
#   takes_em = "Takes_em",
#   effort_em = "Sets_em",
#   effort_total = "Sets_total",  # Old approach
#   family = "poisson",
#   time_varying = FALSE
# )
```

The function automatically detects multi-stream mode and should display:

```
Observer stream: 10 observations
  Total takes: 16
  Total effort: 1000
EM stream: 10 observations
  Total takes: 9
  Total effort: 800
Total fishery effort: 5000
Observed effort: 1800
Unobserved effort: 3200
```

### Plotting results

The plotting functions work the same way:

```{r, fig.pos="placeHere", fig.cap="Estimated bycatch from multi-stream model"}
plot_fitted(fit, xlab = "Year", ylab = "Estimated bycatch", include_points = TRUE)
```

```{r, fig.pos="placeHere", fig.cap="Expanded bycatch estimates (total fishery)"}
plot_expanded(fit, xlab = "Year", ylab = "Total fishery bycatch")
```

### Stream-specific summaries

Get detailed summaries by monitoring stream:

```{r}
stream_summary <- get_stream_summary(fit)
print(stream_summary)
```

This table shows:
- Takes and effort for each stream (Observer, EM, Pooled)
- Coverage percentages
- Estimated total bycatch with credible intervals

### Three-stream example

You can also include a third stream for vessels with both Observer and EM:

```{r data3}
d3 <- data.frame(
  Year = 2010:2019,
  
  # Observer-only stream
  Takes_obs = c(2, 1, 3, 0, 2, 1, 0, 2, 1, 3),
  Sets_obs = rep(80, 10),
  CovRate_obs = rep(16, 10),  # 16% coverage
  
  # EM-only stream
  Takes_em = c(1, 0, 2, 1, 1, 0, 2, 1, 0, 1),
  Sets_em = rep(70, 10),
  CovRate_em = rep(14, 10),   # 14% coverage
  
  # Both Observer and EM
  Takes_both = c(1, 1, 0, 1, 0, 1, 1, 0, 1, 0),
  Sets_both = rep(50, 10),
  CovRate_both = rep(10, 10),  # 10% coverage
  
  # Total fishery effort (optional)
  Sets_total = rep(500, 10)
)
```

```{r, results="hide", message=FALSE, warning=FALSE}
fit3 <- fit_bycatch(
  Takes_obs ~ 1,
  data = d3,
  time = "Year",
  effort = "Sets_obs",
  covrate_obs = "CovRate_obs",
  takes_em = "Takes_em",
  effort_em = "Sets_em",
  covrate_em = "CovRate_em",
  takes_both = "Takes_both",        # Third stream
  effort_both = "Sets_both",
  covrate_both = "CovRate_both",    # Third stream coverage
  family = "poisson",
  time_varying = FALSE
)
```

### Comparing to pooled data

You can verify that multi-stream gives similar results to pooling:

```{r, results="hide", message=FALSE, warning=FALSE}
# Manually pool the data
d_pooled <- d
d_pooled$Takes_pooled <- d$Takes_obs + d$Takes_em
d_pooled$Sets_pooled <- d$Sets_obs + d$Sets_em

d_pooled$CovRate_pooled <- d$CovRate_obs + d$CovRate_em

fit_pooled <- fit_bycatch(
  Takes_pooled ~ 1,
  data = d_pooled,
  time = "Year",
  effort = "Sets_pooled",
  covrate = "CovRate_pooled",  # Add this
  family = "poisson",
  time_varying = FALSE
)
```

Extract and compare the estimated rates:

```{r}
# Multi-stream: lambda_base is rate per unit effort
lambda_multi <- rstan::extract(fit$fitted_model, "lambda_base")$lambda_base
rate_multi <- mean(lambda_multi)

# Single-stream: lambda is expected count, divide by effort
lambda_pooled <- rstan::extract(fit_pooled$fitted_model, "lambda")$lambda
rate_pooled <- mean(lambda_pooled[, 1]) / d_pooled$Sets_pooled[1]

cat("Multi-stream rate:", round(rate_multi, 4), "\n")
cat("Pooled rate:", round(rate_pooled, 4), "\n")
```

The rates should be very similar, confirming that multi-stream mode properly pools information across streams.

### Different distributions

Multi-stream mode works with all distribution families:

```{r, eval=FALSE}
fit_nb <- fit_bycatch(Takes_obs ~ 1, data = d, time = "Year",
  effort = "Sets_obs", covrate_obs = "CovRate_obs",
  takes_em = "Takes_em", effort_em = "Sets_em", covrate_em = "CovRate_em",
  family = "nbinom2", time_varying = FALSE)

fit_hurdle <- fit_bycatch(Takes_obs ~ 1, data = d, time = "Year",
  effort = "Sets_obs", covrate_obs = "CovRate_obs",
  takes_em = "Takes_em", effort_em = "Sets_em", covrate_em = "CovRate_em",
  family = "poisson-hurdle", time_varying = FALSE)
```

### Adding covariates

Covariates work the same way as in single-stream mode:

```{r}
# Add a regulatory change covariate
d$Regulation <- ifelse(d$Year < 2015, 0, 1)
```

```{r, results="hide", message=FALSE, warning=FALSE}
fit_cov <- fit_bycatch(
  Takes_obs ~ Regulation,
  data = d,
  time = "Year",
  effort = "Sets_obs",
  covrate_obs = "CovRate_obs",
  takes_em = "Takes_em",
  effort_em = "Sets_em",
  covrate_em = "CovRate_em",
  family = "poisson",
  time_varying = FALSE
)
```

```{r}
# Extract covariate effects
betas <- rstan::extract(fit_cov$fitted_model)$beta
# beta[,1] = intercept, beta[,2] = Regulation effect
```

### Time-varying effects

You can combine multi-stream with time-varying effects:

```{r, results="hide", message=FALSE, warning=FALSE}
fit_tv <- fit_bycatch(
  Takes_obs ~ 1,
  data = d,
  time = "Year",
  effort = "Sets_obs",
  covrate_obs = "CovRate_obs",
  takes_em = "Takes_em",
  effort_em = "Sets_em",
  covrate_em = "CovRate_em",
  family = "poisson",
  time_varying = TRUE  # Enable time-varying effects
)
```
