---
title: "Multi-Stream Bycatch Estimation"
format: pdf
---

```{r setup, results='hide'}
library(bycatch)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
# Set theme for plots
theme_set(theme_minimal(base_size = 12))

knitr::opts_chunk$set(
  echo = TRUE,       # Show code by default
  message = FALSE,   # Suppress messages from packages
  warning = FALSE    # Suppress warnings
)
```

# Overview

This document presents an overview of modelled bycatch estimates for the following species:

```{r echo = FALSE}

if(file.exists("csem_mmsb_forbayes_wide.csv")) {
  d <- read.csv("csem_mmsb_forbayes_wide.csv")
  spec_names <- unique(d$Species)
  print(spec_names)
}

```

# Methods

We're going to use 2 versions of these models for bycatch estimation

1.  **Multi-stream model**: Combines Observer, EM, and Both monitoring streams
2.  **Observer-only model**: This model is identical to previous reports (Jannot et al.) and based only on Observer data (where available)

## Data Sources

There are up to three data streams for each species

-   **Observer monitoring**: At-sea observer coverage
-   **Electronic monitoring (EM)**: Electronic monitoring (camera-based)
-   **Both**: Trips with both Observer and EM present

## Statistical Models

All models use a Bayesian Poisson GLM framework:

$$
\text{Takes}_i \sim \text{Poisson}(\lambda \times \text{Effort}_i)
$$

where $\lambda$ is the bycatch rate (takes per unit effort).

### Backwards compatibility: 1-stream models

To illustrate the backwards compatibility, we will use any data with an observer present as 'observer' and expand to the unobserved portions of the fleet. Note: the argument `expansion_rate` is used here, and means the same thing as the coverage rate.

```{r fit-models, cache=TRUE, results='hide'}
fit_obs <- list()

for(i in 1:length(spec_names)) {
  # filter by species
  sub <- dplyr::filter(d, Species == spec_names[i])

  # replace all NAs in the data with 0s
  sub <- sub |>
  dplyr::mutate(across(where(is.numeric), ~ tidyr::replace_na(.x, 0)))

  # combined Ret_OBS and Ret_OBS_EM
  sub$Ret_OBS <- sub$Ret_OBS + sub$Ret_OBS_EM
  sub$Takes_OBS <- sub$Takes_OBS + sub$Takes_OBS_EM
  sub$expansion_rate <- 100 * sub$Ret_OBS / (sub$Ret_OBS + sub$Ret_EM)
  
  fit_obs[[i]] <- fit_bycatch(
    Takes_OBS ~ 1,
    data = sub,
    time = "Year",
    effort = "Ret_OBS",
    expansion_rate = "expansion_rate",
    family = "poisson",
    time_varying = FALSE
  )
}
```

Then for each of these models, we can use the package functions to extract any quantities of interest.

```{r eval = FALSE}
get_fitted(fit_obs[[1]])
get_expanded(fit_obs[[1]])
get_total(fit_obs[[1]])
```

or plots can be made with

```{r eval = TRUE}
plot_fitted(fit_obs[[1]]) + 
  ggtitle(spec_names[1])
```

```{r eval = TRUE}
plot_expanded(fit_obs[[1]], show_total = FALSE) + 
  ggtitle(spec_names[1])
```

The expanded estimates here represent the expanded estimates to the unobserved portion of the fleet.  


```{r fit-joined-models, cache=TRUE, eval=TRUE, results='hide'}
fit_joined <- list()
for(i in 1:length(spec_names)) {
  sub <- dplyr::filter(d, Species == spec_names[i])
  
  # replace all NAs in the data with 0s
  sub <- sub |>
  dplyr::mutate(across(where(is.numeric), ~ tidyr::replace_na(.x, 0)))
  orig_retained <- dplyr::select(sub, Year, Retained) |>
    dplyr::rename(raw_retained = Retained)

  # Prior to 2024, 100% coverage means retained = EM + OBS_EM
  sub$Retained_sum <- sub$Ret_EM + sub$Ret_OBS_EM
  
  # 2024 and beyond: use Retained data in original file
  sub <- dplyr::left_join(sub, orig_retained) |>
    dplyr::mutate(Retained = ifelse(Year < 2024, Retained_sum, raw_retained)) |>
    dplyr::select(-Retained_sum, -raw_retained)
  
  fit_joined[[i]] <- fit_bycatch(
  Takes_OBS ~ 1,  # Formula for the observer stream
  data = sub,
  time = "Year",
  # Observer stream (base)
  effort = "Ret_OBS",
  # EM-only stream
  takes_em = "Takes_EM",
  effort_em = "Ret_EM",
  # Both streams present
  takes_both = "Takes_OBS_EM",
  effort_both = "Ret_OBS_EM",
  # Total fishery effort (for expansion)
  effort_total = "Retained",
  # Model settings
  family = "poisson",
  time_varying = FALSE)
}

```

# Results

We can then compare estimates from just using the observer data to the combined approach using any of the above metrics, e.g. 

```{r eval = FALSE, echo=TRUE}
get_total(fit_joined[[1]])
get_expanded(fit_joined[[1]])
get_fitted(fit_joined[[1]])

```

